---
title: "Prosper Loan Exploration by Michael Phillips"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, message=FALSE,warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

loan_df_subset <- read.csv('loan_df_subset.csv')
```

> This report explores a dataset containing approximately 110,000 loans issued by Prosper. Prosper describes itself as America's first "marketplace lending platform", where a person posts to the website the amount of money they are seeking and the reason, and a private investor can loan that amount plus interest if they choose. 

## Univariate Plots

```{r echo=FALSE, message=FALSE,warning=FALSE}
#number of rows and columns
dim(loan_df_subset)
```

```{r echo=FALSE, message=FALSE,warning=FALSE}
str(loan_df_subset)
```

> Our dataset contains 21 variables (plus an index, labeled 'X'), and 113,937 observations.

```{r echo=FALSE, message=FALSE,warning=FALSE}
ggplot(aes(x = EmploymentStatus), data = subset(loan_df_subset, !is.na(EmploymentStatus))) +
    geom_histogram(binwidth = 0.25, stat = 'count', color = 'black', fill = '#F79420') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> As expected, most people are employed. It is surprising some people received loans by not listing any employment status.

```{r echo=FALSE, message=FALSE,warning=FALSE}
ggplot(aes(x = CreditGrade), data = loan_df_subset) +
    geom_histogram(binwidth = 0.25, stat = 'count', color = 'black', fill = '#F79420') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> As soon as I saw this histogram I had to find a reason why more than 80,000 loans were not given a credit grade - the Variable Definitions explain that only listings from pre-2009 were given a Credit Grade.

```{r echo=FALSE, message=FALSE,warning=FALSE}
ggplot(aes(x = IncomeRange), data = loan_df_subset) +
    geom_histogram(binwidth = 0.25, stat = 'count', color = 'black', fill = '#F79420') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> This chart shows a concentration around $50,000, which is very close to the median household income estimated by the Federal Reserve fo 2015. For a bit more detail, lets look at some descriptive statistics for the variable, "StatedMonthlyIncome".

```{r echo=FALSE, message=FALSE,warning=FALSE}
summary(loan_df_subset$StatedMonthlyIncome)
```

> Wow! Someone apparently has a monthly income of almost 2 million dollars and is asking for a loan. Let's look at that entry from the dataframe.

```{r echo=FALSE, message=FALSE,warning=FALSE}
income_outlier <- subset(loan_df_subset, StatedMonthlyIncome > 1000000)
income_outlier[, c("ListingNumber", "EmploymentStatus", "StatedMonthlyIncome"), drop = FALSE]
```

> He or she states that they are self-employed. I suppose its not impossible for this obvious outlier to be a data entry error. Are there any other data points above $500,000?

```{r }
sum(loan_df_subset$StatedMonthlyIncome > 500000)
```

> I want to take a look at the listing for the other point as well.

```{r echo=FALSE, message=FALSE,warning=FALSE}
income_outlier_fivehundred <- subset(loan_df_subset, StatedMonthlyIncome > 500000 & ListingNumber != 560783)
income_outlier_fivehundred[, c("ListingNumber", "EmploymentStatus", "StatedMonthlyIncome"), drop = FALSE]
```

> The other person is also self-employed. Interesting, again, not totally unreasonable.


```{r echo=FALSE, message=FALSE,warning=FALSE}
loan_df_subset$meanCreditScore <- (loan_df_subset$CreditScoreRangeLower + loan_df_subset$CreditScoreRangeUpper) / 2
ggplot(aes(x = meanCreditScore), data = loan_df_subset) +
    geom_histogram(fill = '#F79420') +
    xlab('Mean Credit Score')
    
```

> Most of the Credit Scores are between 425 and 875. Lets zoom in on that region.

```{r echo=FALSE, message=FALSE,warning=FALSE}
ggplot(aes(x = meanCreditScore), data = loan_df_subset) +
    geom_histogram(fill = '#F79420') +
    scale_x_continuous(limits = c(425, 875), breaks = seq(425, 875, 25)) +
    xlab('Mean Credit Score')
```

> We can see the negative skew of this graph, with most people falling within the same 'bucket' of credit scores. I'm curious how many exactly are under 425.

```{r echo=FALSE, message=FALSE,warning=FALSE}
credit_under_fourhundred <- subset(loan_df_subset, meanCreditScore < 425)
nrow(credit_under_fourhundred)

```

> Out of 113,937 possible loans, only 134 fall under a 425 Credit Score.







